---
title: "by_gene_survival"
author: "Gordana"
date: "`r Sys.Date()`"
output: html_document
params: 
  resdir: "results/Subtype"
  ddatadir: "data/derived"
  filenm: "cleaned_gene_dat.Rdata"
  gene: "GUSB"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE, warning=FALSE}
library(survival) # survival models
library(survminer) #plotting survival
library(ggplot2) # plotting
library(dplyr) # data wrangling
library(tibble) # data wrangling
library(table1) # nice summay tables
```


Create results directory if it does not exists
```{r}
if(!dir.exists(paste0("../../",params$resdir))) dir.create(paste0("../../",params$resdir))
```

Load data
```{r}
load(file=paste0("../../",params$ddatadir, "/", params$filenm))
gene_surv_dat <- cdat1 %>% 
    rename(gene = params$gene) %>% #select gene
    mutate_at(c("stage", "mstype", "site"), as.factor) %>% # make these factors
    mutate(stage = relevel(stage, ref = "2"),# make factors and relevel
           gene = scale(gene)) %>% # scale gene
    select(gene, timeint_revised, timelastfu, event, refage_revised,
           bs1, bs2, bs3, bs4, stage, site, mstype)  # select relevant variables

rm(cdat1)
table1( ~ . -bs1 - bs2 - bs3 - bs4 - site  | mstype, data = gene_surv_dat)
```


### Survival analysis

Left truncation is accounted for by removing patients from the risk set until they enrol in the study, as in  Moore (2016, Section 3.5). 


```{r}
my.surv.tmp = with(gene_surv_dat,
                   Surv(time  = timeint_revised/365, # time from diagnosis to study enrolment
                        time2 = timelastfu/365, # time from diagnosis to death/censoring
                        event = event)) # 0 = alive, 1 = dead

plot(survfit(my.surv.tmp~1))
```

Fit model for survival by gene, controlling for age (bs spline), stage, mstype, and stratifying by site.
```{r}
mod_main = coxph( my.surv.tmp ~ bs1 + bs2 + bs3 + bs4 + #bs1 to bs4 are age basis splines
                      stage  + mstype + gene + strata(site),
                    data = gene_surv_dat,
                    na.action = na.exclude)
```

Fit a model as above, adding interaction between gene and mstype
```{r}
mod_int = update(mod_main, ~ . + mstype:gene)
```

Arrange results into a one row data frame
```{r}
res <- summary(mod_main)$conf.int %>% # start with summary of main model
    as.data.frame() %>%  
    transmute(hr = `exp(coef)`,
              hr.L95 = `lower .95`,
              hr.U95 = `upper .95`) %>%  #rename and select variables
    rownames_to_column(var = "pred") %>%
    filter(pred == "gene") %>%  # filter just the gene row
    mutate(gene = params$gene,  # add other things
           p.gene = anova(mod_main)[ "gene", "Pr(>|Chi|)"],
           p.ms.gene.int = anova(mod_int)[ "mstype:gene", "Pr(>|Chi|)" ],
           n = mod_main$n) %>% 
    relocate(gene) %>% 
    select(-pred)

res
```

### Violin plot

```{r}
ggplot( gene_surv_dat, aes( mstype, gene ) ) + 
    xlab("molecular subtype") + ylab("normalized expression") + theme_bw() +
    geom_violin(aes(fill=mstype)) + 
    geom_point(size=.05) +
    geom_jitter(width = 0.05, height=0.0, alpha=.2) + 
    labs(title=params$gene) + guides(fill=FALSE) +
    geom_boxplot(width=0.05, notch=TRUE, outlier.shape=NA) +
    stat_compare_means(symnum.args = )
```


```{r, fig.width=5, fig.height=5}

# calculate gene quantiles
gene_surv_dat <- gene_surv_dat %>% 
  mutate(qgene = cut(gene, quantile(gene,seq(0, 1, 0.20)), include.lowest = TRUE),
         qgene = factor(qgene, levels = levels(qgene),
                        labels = c("Q1", "Q2", "Q3", "Q4", "Q5"))) # relabel


# calculate and format p-value (for left truncated can't use default log rank, but can use cox version)
log_rank = summary(coxph(my.surv.tmp   ~ qgene ,
                         data = gene_surv_dat))$sctest # log rank test
pval = ifelse(log_rank[3]<0.0001, "p < 0.0001", paste("p =", round(log_rank[3], 4))) 


# plot survival curves
# fit model
qgene_survmod <- survfit(my.surv.tmp ~ qgene, data=gene_surv_dat)

qgene_surv_plot = ggsurvplot(qgene_survmod, data=gene_surv_dat, submain=params$gene, 
                      conf.int = FALSE, 
                      risk.table = TRUE, 
                      risk.table.y.text.col = TRUE,
                      risk.table.height = 0.3,
                      surv.median.line = "hv",  # add the median survival pointer.
                      xlab="Time (years)",
                      ggtheme = theme_bw(), legend="bottom",
                      legend.labs = c("Q1", "Q2", "Q3", "Q4", "Q5"),
                      palette = c( "#99CCCC", "#0099CC", "#6666FF", "#0033CC", "#663366" ) ) 
# the above palette is not that colour blind friendly

# add p-value text
qgene_surv_plot$plot <- qgene_surv_plot$plot+ 
              ggplot2::annotate("text", 
                                x = 9, y = 0.95,
                                label = pval, size = 4)

qgene_surv_plot

```


### References

Moore, Dirk F. Applied survival analysis using R. Vol. 473. Cham: Springer, 2016. [web link](https://xsliulab.github.io/Workshop/2021/week3/survival-analysis-book.pdf)

